Transient dynamics of granular piles: continuum v discrete modelling

### 2016-07-14 paper.tex
Removed MPM initial stress and particles

% Template for Elsevier CRC journal article
% version 1.2 dated 09 May 2011

% This file (c) 2009-2011 Elsevier Ltd.  Modifications may be freely made,
% provided the edited file is saved under a different name

% This file contains modifications for Procedia Engineering

% Changes since version 1.1
% - added "procedia" option compliant with ecrc.sty version 1.2a
%   (makes the layout approximately the same as the Word CRC template)
% - added example for generating copyright line in abstract

%-----------------------------------------------------------------------------------

%% This template uses the elsarticle.cls document class and the extension package ecrc.sty
%% For full documentation on usage of elsarticle.cls, consult the documentation "elsdoc.pdf"
%% Further resources available at http://www.elsevier.com/latex

%-----------------------------------------------------------------------------------

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                          %%
%% Important note on usage                                  %%
%% -----------------------                                  %%
%% This file should normally be compiled with PDFLaTeX      %%
%% Using standard LaTeX should work but may produce clashes %%
%%                                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% The '3p' and 'times' class options of elsarticle are used for Elsevier CRC
%% The 'procedia' option causes ecrc to approximate to the Word template
\documentclass[3p,times,procedia,number]{elsarticle}
\flushbottom

%% The `ecrc' package must be called to make the CRC functionality available
\usepackage{ecrc}
%\usepackage{amsmath}

%% The ecrc package defines commands needed for running heads and logos.
%% For running heads, you can set the journal name, the volume, the starting page and the authors

%% set the volume if you know. Otherwise `00'
\volume{00}

@@

%% Give the author list to appear in the running head
%% Example \runauth{C.V. Radhakrishnan et al.}
\runauth{K. Kumar, J.-Y. Delenne and K. Soga}

%% The choice of journal logo is determined by the \jid and \jnltitlelogo commands.
%% A user-supplied logo with the name <\jid>logo.pdf will be inserted if present.
%% e.g. if \jid{yspmi} the system will look for a file yspmilogo.pdf
%% Otherwise the content of \jnltitlelogo will be set between horizontal lines as a default logo

@@

%% Give a short journal name for the dummy logo (if needed)
%\jnltitlelogo{Procedia Engineering}

%% Hereafter the template follows `elsarticle'.
%% For more details see the existing template files elsarticle-template-harv.tex and elsarticle-template-num.tex.

%% Elsevier CRC generally uses a numbered reference style
%% For this, the conventions of elsarticle-template-num.tex should be followed (included below)
%% If using BibTeX, use the style file elsarticle-num.bst

%% End of ecrc-specific commands
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{cleveref}
\graphicspath{figs/}
% ***************************** Math and SI Units ******************************
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[per=slash]{siunitx} % use this package module for SI units
\usepackage{flexisym}
\usepackage{subcaption}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers after \end{frontmatter}.
%% \usepackage{lineno}

%% natbib.sty is loaded by default. However, natbib options can be
%% provided with \biboptions{...} command. Following options are
%% valid:

%%   round  -  round parentheses are used (default)
%%   square -  square brackets are used   [option]
%%   curly  -  curly braces are used      {option}
%%   angle  -  angle brackets are used    <option>
%%   semicolon  -  multiple citations separated by semi-colon
%%   colon  - same as semicolon, an earlier confusion
%%   comma  -  separated by comma
%%   numbers-  selects numerical citations
%%   super  -  numerical citations as superscripts
%%   sort   -  sorts multiple citations according to order in ref. list
%%   sort&compress   -  like sort, but also compresses numerical citations
%%   compress - compresses without sorting
%%
%\biboptions{authoryear}

@@

% if you have landscape tables
\usepackage[figuresright]{rotating}
%\usepackage{harvard}
% put your own definitions here:x
%   \newcommand{\cZ}{\cal{Z}}
%   \newtheorem{def}{Definition}[section]
%   ...

% add words to TeX's hyphenation exception list
%\hyphenation{author another created financial paper re-commend-ed Post-Script}

% declarations for front matter

@@

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for the associated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for the associated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for the associated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%%
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

\dochead{1st International Conference on the Material Point Method, MPM 2017}
%% Use \dochead if there is an article header, e.g. \dochead{Short communication}
%% \dochead can also be used to include a conference title, if directed by the editors
%% e.g. \dochead{17th International Conference on Dynamical Processes in Excited States of Solids}

@@

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{<author name>}
%% \address[label1]{<address>}
%% \address[label2]{<address>}

@@

%% keywords here, in the form: keyword \sep keyword

@@

%\correspondingauthor[*]{Corresponding author. Tel.: +0-000-000-0000 ; fax: +0-000-000-0000.}
\email{kks32@cam.ac.uk}

%%
%% Start line numbering here if you want
%%
% \linenumbers

@@

Granular flows have been studied in laboratory experiments in different 
geometries such as tilted piles leading to slope failure and surface avalanches 
\citep{Legros2002, Iverson1997} or by considering vertical columns 
of grains collapsing and spreading under their own weight \citep{Lajeunesse2004, 
Lajeunesse2005}. The dynamics observed in the experiments is often nontrivial 
in the sense that the final configurations after the dissipation of the whole kinetic energy  
can not be readily predicted by means of simple laws based on the Mohr-Coulomb nature 
of the material. For example, in collapsing columns, the runout distance 
is found to obey a power-law dependence upon the initial aspect ratio of the 
column.

The observed nontrivial transient dynamics is often correctly 
reproduced by the discrete element method (DEM), which provides 
a powerful tool for the grain-scale analysis of the 
trigger and its subsequent dynamics \citep{Staron2005, Staron2009}. 
However, even in simplified geometries such as those 
investigated in the experiments, the DEM suffers from a serious 
short-coming in the number of particles that can be simulated in 
a reasonnable time. This is a critical issue when more complex geometries or 
long-time granular processes are considered, or when particle size 
distributions are broad. For this reason, most numerical studies 
are performed in 2D or simple particles shapes and size distributions 
are considered.

@@

In this paper, we are concerned with the ability of the MPM, as a continuum 
approach, to reproduce the evolution of a granular pile 
under its own weight or when destabilized by energy input. In particular, 
a central issue is wheather power-law dependence of the runout distance 
and timewith respect to the initial geometry or energy can be reproduced 
by a simple prescription of the Mohr-Coulomb plastic behavior within a  
MPM code. We therefore perform extensive simulations by varying continuously 
different input parameters and compare the data with those obtained from 
DEM simulations of the same system. We compare in detail the evolution of 
the profile of the pile and its total kinetic energy between the two methods 
and for differerent initial states. As we shall see, the MPM can successfully simulate  
the transient evolution with a single input parameter, namely the 
internal angle of friction. This opens the way to the simulation of 
geological-scale flows on complex topographies.

@@

For grain-scale simulations, classical DEM and Contact Dynamics approaches were 
used. This research was done in collaboration with Patrick Mutabaruka, 
University of Montpellier, who performed Contact Dynamics (CD) simulations that 
are presented in this section. A detailed description of the Contact Dynamics 
method can be found in \cite{Moreau1993,Jean1999,Radjai2009,Radjai2011}. 
The CD method is based on implicit time integration of the equations of motion 
and a non-smooth formulation of mutual exclusion and dry friction between 
particles. The CD method requires no elastic repulsive potential and no 
smoothing of the Coulomb friction law for the determination of forces. 
For this reason, the simulations can be performed with large time steps 
compared to discrete element simulations. The unknown variables are particle 
velocities and contact forces, which are calculated at each time step by taking 
into account the conservation of momenta and the constraints due to mutual 
exclusion between particles and the Coulomb friction. An iterative 
algorithm based on a non-linear Gauss-Seidel scheme is used. The only 
contact parameters within the CD method are the friction coefficient $\mu$, the 
normal restitution coefficient $\varepsilon_n$ and the tangential restitution 
coefficient $\varepsilon_t$ between grains.

In MPM simulations, the material point spacing is adopted to be the same as the 
mean grain diameter in DEM. A mesh size of 0.0125m is adopted with 25 material 
points per cell. The effect of mesh size and the number of material points per 
cell is investigated in~\cref{sec:MPM_points_per_cell}. The initial 
configuration of the slope in MPM is shown in~\cref{fig:MPM_Initial_PIC_Slope}. 
Frictional boundary conditions are applied on the left and the bottom 
boundaries by applying constraints to the nodal acceleration. The initial 
vertical stress of the granular pile in equilibrium, before the horizontal 
excitation, in the MPM simulation is shown in~\cref{fig:MPM_Stress_Slope}. The 
distribution of the initial gradient horizontal excitation energy of 50~\si{J} 
on the granular pile is shown in~\cref{fig:MPM_Velocity_Slope}. The 
Mohr-Coulomb model with no dilation is used to simulate the continuum behaviour 
of the granular pile. Periodic shear tests using CD 
(\cref{fig:Sxy_vs_Syy_Slope}), reveals a macroscopic friction coefficient of 
0.22. The evolution of inertial number with friction is presented 
in~\cref{fig:mu_vs_I}.

\begin{figure}[tbhp]
  \centering
  \begin{subfigure}[t]{0.75\textwidth}
    \centering
    \includegraphics[width=0.95\textwidth]{figs/MPM_Initial_PIC_Slope}
    \caption{Initial configuration of material points per cell. 25 material points 
    per cell. Cell size of 0.0125m.}
    \label{fig:MPM_Initial_PIC_Slope}
  \end{subfigure} \\
  %
  \begin{subfigure}[t]{0.95\textwidth}
    \centering
    \includegraphics[width=0.95\textwidth]{figs/MPM_Stress_Slope}
    \caption{Initial stress in MPM}
    \label{fig:MPM_Stress_Slope}
  \end{subfigure} \\
  %
  \begin{subfigure}[t]{0.95\textwidth}
    \centering
    \includegraphics[width=0.95\textwidth]{figs/MPM_Velocity_Slope}
    \caption{Initial horizontal velocity for a pile subjected to a horizontal 
    velocity of 50J.}
    \label{fig:MPM_Velocity_Slope}
  \end{subfigure}
  \caption{Initial configuration for the MPM simulation of a pile subjected to 
  horizontal velocities.}
  \label{fig:MPM_Slope_setup}
\end{figure}

\begin{figure}[tbhp]
  \centering
  \begin{subfigure}[t]{0.475\textwidth}
    \includegraphics[width=0.95\textwidth]{figs/Sxy_vs_Syy_Slope}
    \caption{Evaluation of the critical state friction angle.}
    \label{fig:Sxy_vs_Syy_Slope}
  \end{subfigure}
  %
  \begin{subfigure}[t]{0.475\textwidth}
    \includegraphics[width=0.95\textwidth]{figs/mu_vs_I}
    \caption{Evolution of Inertial number with friction $\mu$.}
    \label{fig:mu_vs_I}
  \end{subfigure}
  \caption{Periodic shear test using CD~\citep{Mutabaruka2013}.}
  \label{fig:Shear_Test_Slope}
\end{figure}

>>>>

\documentclass[3p,times,procedia,number]{elsarticle}
\flushbottom

%% The `ecrc' package must be called to make the CRC functionality available
\usepackage{ecrc}
%\usepackage{amsmath}
\volume{00}

@@

%% Give the author list to appear in the running head
\runauth{K. Kumar, J.-Y. Delenne and K. Soga}

@@

%% End of ecrc-specific commands
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{cleveref}
\graphicspath{figs/}
% ***************************** Math and SI Units ******************************
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[per=slash]{siunitx} % use this package module for SI units
\usepackage{flexisym}
\usepackage{subcaption}

@@

% if you have landscape tables
\usepackage[figuresright]{rotating}

@@

\dochead{1st International Conference on the Material Point Method, MPM 2017}

@@

@@

@@

\email{kks32@cam.ac.uk}

%% Start line numbering here if you want
% \linenumbers

@@

Granular flows have been studied in laboratory experiments in different 
geometries such as tilted piles leading to slope failure and surface
avalanches~\citep{Legros2002, Iverson1997} or by considering vertical columns 
of grains collapsing and spreading under their own
weight~\citep{Lajeunesse2004, Lajeunesse2005}. The dynamics observed in the
experiments is often nontrivial in the sense that the final configurations
after the dissipation of the whole kinetic energy can not be readily predicted
by means of simple laws based on the Mohr-Coulomb nature of the material. For
example, in collapsing columns, the runout distance is found to obey a
power-law dependence upon the initial aspect ratio of the column.

The observed nontrivial transient dynamics is often correctly reproduced by the
discrete element method (DEM), which provides a powerful tool for the
grain-scale analysis of the trigger and its subsequent
dynamics~\citep{Staron2005, Staron2009}. However, even in simplified geometries
such asthose investigated in the experiments, the DEM suffers from a
seriousshort-coming in the number of particles that can be simulated in a
reasonnable time. This is a critical issue when more complex geometries or
long-time granular processes are considered, or when particle size
distributions are broad. For this reason, most numerical studies are performed
in 2D or simple particles shapes and size distributions are considered.

@@

In this paper, we investigate the ability of the MPM, as a continuum
approach, to reproduce the evolution of a granular pile under its own weight or
when destabilized by energy input. In particular, a central issue is wheather
power-law dependence of the runout distance and timewith respect to the initial
geometry or energy can be reproduced by a simple prescription of the
Mohr-Coulomb plastic behavior within a  MPM code. We therefore perform
extensive simulations by varying continuously different input parameters and
compare the data with those obtained from DEM simulations of the same system.
We compare in detail the evolution of the profile of the pile and its total
kinetic energy between the two methods and for differerent initial states. As
we shall see, the MPM can successfully simulate  the transient evolution with a
single input parameter, namely the internal angle of friction. This opens the
way to the simulation of geological-scale flows on complex topographies.

@@

For grain-scale simulations, classical DEM approach was used.

In MPM simulations, the material point spacing is adopted to be the same as the 
mean grain diameter in DEM. A mesh size of 0.0125m is adopted with 25 material 
points per cell. The effect of mesh size and the number of material points per 
cell was investigated by~\citep{soundararajan2015}. The initial 
configuration of the slope in MPM is shown in~\cref{fig:mpm_slope_setup}. 
Frictional boundary conditions are applied on the left and the bottom 
boundaries by applying constraints to the nodal acceleration. Mohr-Coulomb
model with no dilation is used to simulate the continuum behaviour of the
granular pile. Periodic shear tests in DEM reveals a macroscopic friction 
coefficient of 0.22.

\begin{figure}[tbhp]
  \centering
  \includegraphics[width=0.95\textwidth]{figs/mpm_velocity_slope}
  \caption{Initial configuration for the MPM simulation of a pile subjected to 
  horizontal velocity of 50J.}
  \label{fig:mpm_slope_setup}
\end{figure}
